\morus: Biases in the Stream

### 2018-05-10 morusAC_01_Introduction.tex
minor Section 2 and rounds vs. steps

\subsubsection*{Related Work.}
In the \MORUS submission document, the designers discuss the security of \MORUS against
several attacks, including algebraic, differential, and guess-and-determine attacks.
The main focus is on differential properties, and not many details are given for other attack vectors.
In third-party analysis,
Mileva et~al.~\cite{balkancryptsecMilevaDV15} propose a distinguisher in the nonce-reuse setting and practically evaluate the differential behaviour of toy variants of \MORUS.
Dwivedi et~al.~\cite{cryptoeprint:2016:1053} discuss the applicability of SAT solvers for state recovery, but the resulting complexity of $2^{370}$ for \MORUS[640] is well beyond the security claim.
Dwivedi et~al.~\cite{secryptDwivediMW17} also propose key-recovery attacks for \MORUS[1280] if initialization is reduced to 3.6 out of 16 rounds, and discuss the security of \MORUS against internal differentials and rotational cryptanalysis.
Salam et~al.~\cite{trustcomSalamSBDPW17} apply cube attacks to obtain distinguishers for up to 5 out of 16 rounds of the initialization of \MORUS[1280] with negligible complexity.
Additionally, 
Kales et~al.~\cite{cryptoeprint:2017:1137} and
Vaudenay and Viz\'{a}r \cite{cryptoeprint:2017:1147}
independently propose state-recovery and forgery attacks on \MORUS in a nonce-misuse setting with negligible data and time complexities.
Finally, a keystream bias of a similar nature as our main attack was uncovered by Minaud \cite{sacryptMinaud14} on the authenticated cipher \cipher{AEGIS} \cite{AEGIS,sacryptWuP13}, another CAESAR finalist which shares the same overall structure as \MORUS, but uses a very different, \cipher{AES}-based state update function.

@@

In addition to this attack on full \MORUS, we also analyze the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
While this analysis does not threaten full \MORUS, it complements the main result to provide a better understanding of the \MORUS design approach.
We present a forgery attack for round-reduced \MORUS[1280] with success probability $2^{-88}$ for a 128-bit tag if the finalization is reduced to 3 out of 10 rounds. This nonce-respecting attack is based on a differential analysis of the padding rule.
The second result targets a round-reduced initialization with 10 out of 16 rounds, and extends a state-recovery attack (e.g., in a nonce-misuse setting) into a key-recovery attack.

\subsubsection*{Outline.}
This paper is organized as follows.
We first provide a brief description of \MORUS in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \MiniMORUS, an abstraction of \MORUS based on a certain class of rotational symmetries.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a bias in the keystream over 5 rounds,
and discuss the implications of our observation for the security of \MORUS in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \Cref{sec/Conclusion}.

>>>>

\subsubsection*{Related Work.}
In the \MORUS submission document, the designers discuss the security of \MORUS against
several attacks, including algebraic, differential, and guess-and-determine attacks.
The main focus is on differential properties, and not many details are given for other attack vectors.
In third-party analysis,
Mileva et~al.~\cite{balkancryptsecMilevaDV15} propose a distinguisher in the nonce-reuse setting and practically evaluate the differential behaviour of toy variants of \MORUS.
Dwivedi et~al.~\cite{cryptoeprint:2016:1053} discuss the applicability of SAT solvers for state recovery, but the resulting complexity of $2^{370}$ for \MORUS[640] is well beyond the security claim.
Dwivedi et~al.~\cite{secryptDwivediMW17} also propose key-recovery attacks for \MORUS[1280] if initialization is reduced to 3.6 out of 16 steps, and discuss the security of \MORUS against internal differentials and rotational cryptanalysis.
Salam et~al.~\cite{trustcomSalamSBDPW17} apply cube attacks to obtain distinguishers for up to 5 out of 16 steps of the initialization of \MORUS[1280] with negligible complexity.
Additionally, 
Kales et~al.~\cite{cryptoeprint:2017:1137} and
Vaudenay and Viz\'{a}r \cite{cryptoeprint:2017:1147}
independently propose state-recovery and forgery attacks on \MORUS in a nonce-misuse setting with negligible data and time complexities.
Finally, a keystream bias of a similar nature as our main attack was uncovered by Minaud \cite{sacryptMinaud14} on the authenticated cipher \cipher{AEGIS} \cite{AEGIS,sacryptWuP13}, another CAESAR finalist which shares the same overall structure as \MORUS, but uses a very different, \cipher{AES}-based state update function.

@@

In addition to this attack on full \MORUS, we also analyze the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
While this analysis does not threaten full \MORUS, it complements the main result to provide a better understanding of the \MORUS design approach.
We present a forgery attack for round-reduced \MORUS[1280] with success probability $2^{-88}$ for a 128-bit tag if the finalization is reduced to 3 out of 10 steps. This nonce-respecting attack is based on a differential analysis of the padding rule.
The second result targets a round-reduced initialization with 10 out of 16 steps, and extends a state-recovery attack (e.g., in a nonce-misuse setting) into a key-recovery attack.

\subsubsection*{Outline.}
This paper is organized as follows.
We first provide a brief description of \MORUS in \Cref{sec/Preliminaries}.
In \Cref{sec/introminimorus}, we introduce \MiniMORUS, an abstraction of \MORUS based on a certain class of rotational symmetries.
We analyze this simplified scheme in \Cref{sec/minitrails} and provide a ciphertext-only linear approximation with a weight of 16.
We then extend our result to the full scheme in \Cref{sec/fulltrails}, showing a bias in the keystream over 5 steps,
and discuss the implications of our observation for the security of \MORUS in \Cref{sec/Discussion}.
In \Cref{sec/IniFin}, we present our results on the security of \MORUS with round-reduced initialization (in a nonce-misuse setting) or finalization.
Finally, we conclude in \Cref{sec/Conclusion}.

### 2018-05-10 morusAC_02_Preliminaries.tex
minor Section 2 and rounds vs. steps

\MORUS \cite{MORUS} is a family of authenticated ciphers with two internal state sizes: 640 and 1280 bits.
Each member is called \MORUS[640] and \MORUS[1280], respectively.

An instance of \MORUS is parametrized by a secret key. During encryption, it takes as input a plaintext message and possibly some associated data, and outputs a ciphertext together with an authentication tag.
In the recommended parameters, \MORUS[640] supports 128-bit keys and \MORUS[1280] supports both 128-bit and 256-bit keys. The tag size is 128 bits or shorter. The designers strongly recommend using a 128-bit tag.

@@

The internal state of \MORUS is composed of five registers. Let $q$ be the register size, where $q = 128$ for \MORUS[640] and $q=256$ for \MORUS[1280]. Let $S_i, i \in\{0,1,2,3,4\}$ be a register. Then the internal state of \MORUS may be represented as $S_0\|S_1\|S_2\|S_3\|S_4$.
Registers are themselves divided into four $q/4$-bit \emph{words}.
Throughout the paper, we denote the word size by $w = q/4$, i.e. $w=32$ for \MORUS[640] and $w=64$ for \MORUS[1280].

>>>>

\MORUS is a family of authenticated ciphers designed by Wu and Huang \cite{MORUS}.
In this section, we provide a brief description of \MORUS and introduce the
notation for linear approximations.

@@

An instance of \MORUS is parametrized by a secret key $K$. During encryption, it takes as input a plaintext message $M$, a nonce $N$, and possibly some associated data $A$, and outputs a ciphertext $C$ together with an authentication tag $T$.

The \MORUS family supports two internal state sizes: 640 and 1280 bits,
referred to as \MORUS[640] and \MORUS[1280], respectively.
Three parameter sets are recommended: \MORUS[640] supports 128-bit keys and \MORUS[1280] supports either 128-bit or 256-bit keys. The tag size is 128 bits or shorter. The designers strongly recommend using a 128-bit tag.

The internal state of \MORUS is composed of five $q$-bit \emph{registers} $S_i$, $i \in\{0,1,2,3,4\}$, where $q = 128$ for \MORUS[640] and $q=256$ for \MORUS[1280]. The internal state of \MORUS may be represented as $S_0\|S_1\|S_2\|S_3\|S_4$.
Registers are themselves divided into four $q/4$-bit \emph{words}.
Throughout the paper, we denote the word size by $w = q/4$, i.e., $w=32$ for \MORUS[640] and $w=64$ for \MORUS[1280].
